---
title: "Abortion Trends in the United States"
author: Laura Edwards 
date: 7 July 2022
output: 
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    math: mathjax 
---

# Intro
Hi! Today we are taking a look at the most recent CDC Abortion Surveillance Data, from 2019 in conjunction with the 2020 U.S. Census Data to visualize trends in abortion rates in the United States.

First, you need to install and load the following packages:

```{r, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
library(tidyverse)
library(purrr)
library(rgdal)
library(sf)
library(leaflet)
library(tidygeocoder)
library(tidycensus)
#Package gives polygons for US States in data frame called us_states.
library(spData)
```

Now, we need to read the 2019 CDC Abortion Surveillance Data into our session. 
```{r}
abortions_usa <- read.csv("data/abortions-distributed-by-state-2019.csv", skip = 1)
```

# Cleaning CDC Abortion Data
For the purpose of this analysis, we will be considering only abortions that occur within each state, regardless of if abortion patients come from other states. We need to select only the rows and columns from the CDC data table that are relevant to this purpose. 

```{r}
#Select relevant rows and columns
abortions_usa_1 <- abortions_usa[53, 2:53]
#Rename column to accurate, descriptive name
abortions_usa_1 <- rename(abortions_usa_1, Number_of_Abortions = State.Area)
#Pivot data table longer for more sensible information presentation and to prep for palette creation, which creates palettes based on a column, not a row, of values
abortions_usa_2 <- pivot_longer(abortions_usa_1, cols = 1:52, names_to = "State", values_to = "Number_of_Abortions_In_State")
 #Select for relevant rows
abortions_usa_2 <- abortions_usa_2[2:52, ]
#Transform Number_of_Abortions_In_State column from "character" class to "numeric" so that the colorNumeric palette can determine the range of the numbers and assign to color palette
abortions_usa_2$Number_of_Abortions_In_State <- as.numeric(abortions_usa_2$Number_of_Abortions_In_State) 
```
Now, let's make a palette for the Number of Abortions column that we can use on our map later. 
```{r}
palette_abortions <- colorNumeric(palette = "YlOrRd", 
                                  domain = abortions_usa_2$Number_of_Abortions_In_State)
```
In order to create a map, we will need to join the information from the data frame us_states from the spData package, which holds our geometry for the US states and the information from abortions_usa_2, which holds information about abortion numbers.

In order for this to occur, we will need so standardize the format of the names for the states between us_states and abortions_usa_2. 
```{r}
#Removed excess periods from certain states' names
abortions_usa_2$State <- gsub("\\..$", "", abortions_usa_2$State)

#Removed excess periods from certain states' names
abortions_usa_2$State <- gsub("\\.$", "", abortions_usa_2$State)

#Replaced '..." with "_" in certain states
abortions_usa_2$State <- gsub("[...]", "_", abortions_usa_2$State) 

#Renamed New York City and State entry to New York
abortions_usa_2$State[33] <- "New_York"

#Replace every " " with a "_" in us_states data frame to match with the name format for states in abortions_usa_2
us_states$NAME <- gsub(" ", "_", us_states$NAME) 
```
We need to join us_states and abortions_usa_2 by state so palette can be correctly applied on our map.
```{r}
#Join US shapefiles and Abortions data 
us_states_abortions_joined <- inner_join(us_states, abortions_usa_2, by = c("NAME" = "State"))
#Make palette for Number of Abortions in State
palette_abortions <- colorNumeric(palette = "YlOrRd", 
                                  domain = us_states_abortions_joined$Number_of_Abortions_In_State)
```
# Map 1: Number of Abortions By State
```{r, warning = FALSE, message = FALSE}
#Create map
leaflet() %>%
  setView(lat = 40.730610, lng = -98.5556, zoom = 2) %>% 
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(data = us_states_abortions_joined, 
              weight = 1,
              fillColor = ~palette_abortions(Number_of_Abortions_In_State),
              fillOpacity = 1)
```
Normalizing data is important. By calculating an abortion rate and creating a color palette based upon rate over total abortions in the state, we can get a more accurate picture of abortion trends in the United States.  

In order to calculate abortion rate, we need to collect population data for each state. We will obtain this data from the 2020 U.S. Census.
```{r results='hide', fig.keep='all', message = FALSE}
#Import 2020 state population data from U.S. Census
census_states <- get_decennial(geography = "state", variables = "P2_005N", year = 2020, geometry = TRUE, summary_var = "P1_001N", show_call = TRUE)
#Fix names to underscores to prepare for joining with abortions_usa_2
census_states$NAME <- gsub(" ", "_", census_states$NAME)
```
Now, join census_states to abortions_usa_2 so that we can use the mutate() function from the dplyr package within tidyverse to create a new column which will calculate an abortion rate for each state.
```{r}
#Join relevant columns of census_states to abortions_usa_2
census_abortions_joined <- census_states |> 
  select(NAME, summary_value) |>
  inner_join(abortions_usa_2, 
             by = c("NAME" = "State"))
#Create new column in census_abortions_joined that calculates abortions per capita for each state
census_abortions_joined <- census_abortions_joined |> 
  mutate(abortions_per_capita = 
           Number_of_Abortions_In_State / summary_value) 
#Make palette for Map 2
palette_abortions_per_capita <- colorNumeric(
                                  palette = "YlOrRd", 
                                  domain = census_abortions_joined$abortions_per_capita)
        
```
# Map 2: Abortions Per Capita in the United States
```{r, warning = FALSE}
#Create map
leaflet() %>%
  setView(lat = 40.730610, lng = -98.5556, zoom = 2) %>% 
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(data = census_abortions_joined, 
              weight = 1,
              fillColor = ~palette_abortions_per_capita(abortions_per_capita),
              fillOpacity = 1)
```
# Examining the States in with Top 10 Abortion Rates and Political Leaning

Let's read in the Red_Blue_States.csv. This data table was created based off of [analysis from FiveThirtyEight](https://fivethirtyeight.com/features/how-red-or-blue-is-your-state-your-congressional-district/), which labelled states as Red or Blue based off of the state's voting habits compared to the country's overall.  
```{r}
#Load data about political lean
red_blue_states <- read.csv("data/Red_Blue_States.csv") 
#Join information on state political leanings with state abortions information
census_abortions_party <- inner_join(census_abortions_joined, red_blue_states, by = c("NAME" = "State"))
```
Let's select the states with the top 10 abortion rates and visualize the distribution of political leanings on a bar graph.
```{r}
#Create data frame for states with the top 10 abortions per capita
census_abortions_10 <- st_drop_geometry(census_abortions_party) |>
  arrange(desc(abortions_per_capita)) |> 
  head(10)
#Make bar graph
ggplot(census_abortions_10, aes(x = Political.Lean, fill = Political.Lean)) + 
  geom_bar() +
  labs(title = "Political Leanings of States with Top 10 Abortion Rates in \n the United States as of 2019",
       x = "Political Lean", 
       y = "Number of States",) +
  scale_fill_discrete(name = "Political Lean") +
  theme_minimal() 

```





